# 6 Multicategory Logit Models {#Multicategory}

## 6.1 Baseline-Category Logit Models for Nominal Responses

### 6.1.1 Baseline-Category Logits

$$\mathrm{log}\left(\frac{\pi_j}{\pi_c}\right),\ j = 1, \dots,\ c-1.$$

\begin{equation} 
  \mathrm{log}(\frac{\pi_j}{\pi_c}) = \alpha_j + \beta_j x,\ j = 1, \dots, c-1 
  (\#eq:eq61)
\end{equation} 

\begin{equation} 
  \mathrm{log}(\frac{\pi_1}{\pi_2}) = \mathrm{log}\left(\frac{\pi_1/\pi_3}{\pi_2/\pi_3}\right) = \mathrm{log}\left(\pi_1/\pi_3\right) - \mathrm{log}\left(\pi_2/\pi_3\right) \\ 
  = (\alpha_1 + \beta_1 x) -  (\alpha_2 + \beta_2 x) \\
  = (\alpha_1 -\alpha_2) +  (\beta_1 - \beta_2) x 
  (\#eq:eq62)
\end{equation} 

\begin{equation} 
  \mathrm{log}(\frac{\pi_j}{\pi_c}) = \alpha_j + \beta_{j1} x_1 + \beta_{j2} x_2 + \dots + \beta_{jp} x_p \ j = 1, \dots, c-1 
  (\#eq:eq63)
\end{equation} 

### 6.1.2 Example: What Do Alligators Eat?

```{r, 6.1.2a}
library(tidyverse)
Gators <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Alligators.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Gators %>% 
  filter(row_number() %in% c(1, 2, n())) 

library(VGAM)  # package for multivariate GLMs, such s multinomial models 
fit <- vglm(y ~ x, family = multinomial, data = Gators)  # vglm = vector GLM
summary(fit)
```

$$\mathrm{log}(\hat\pi_1/\hat\pi_3) = 1.618 - 0.110x, \\
\mathrm{log}(\hat\pi_1/\hat\pi_2) = 5.697 - 2.465x. $$

$$\mathrm{log}(\hat\pi_1 / \hat\pi_2) = (1.618 - 5.697) + [- 0.110 - (- 2.465)]x = -4.080 + 2.355x.$$

```{r 6.1.2b}
fit2 <- vglm(y ~ x, family = multinomial(refLevel = 2), data = Gators) 
summary(fit2)
confint(fit2, method = "profile")
```

```{r 6.1.2c}
fit0 <- vglm(y ~ 1, family=multinomial, data = Gators)  # null model
deviance(fit0)  # deviance for working model is 98.3412

VGAM::lrtest(fit, fit0)  # rltest function available in VGAM package for LR tests
# .00022 deviance diff = 2(log-like. diff.)
```

### 6.1.3 Estimating Response Probabilities
The response probabilities relate to the model parameters by
$$\pi_j = \frac{e^{\alpha_j + \beta_{j1} x_1 + \beta_{j2} x_2 + \dots + \beta_{jp} x_p}}
{\sum_{h=1}^c e^{\alpha_h + \beta_{h1} x_1 + \beta_{h2} x_2 + \dots + \beta_{hp} x_p}},\ j = 1, \dots , c.$$

```{r}
fitted(fit) %>% 
  data.frame(.) %>% 
  filter(row_number() %in% c(1, 2, n())) 
# obs 59 is alligator of length 3.89 meters
```

```{r}

library(ggthemes)

library(RColorBrewer)
colors <- brewer.pal(n = 4, name = "Dark2")

fitted(fit) %>% 
  data.frame(.) %>% 
  bind_cols(length = Gators$x) %>% 
  rename("Fish" = F, "Invertebrates" = I, "Other" = O) %>% 
  pivot_longer(cols = c(Fish, Invertebrates, Other))  %>% 
  ggplot() + 
  geom_line(aes(x = length, y = value, color = name)) +
  scale_colour_manual(values=colors)+
  ylim(0, 1) + 
  xlim(1, 4) +
  xlab("Length of Alligator") +
  ylab("Predicted Probability") +
  theme_few() +
  annotate(geom = "segment", x = 2.8, y = 0.65, xend = 2.60, yend = .71, 
           arrow = arrow(length = unit(2, "mm")) , color = colors[1]) +
  annotate(geom = "text", x = 2.825, y = 0.65, label = "Fish", 
           hjust = "left", color = colors[1]) +
  
  annotate(geom = "segment", x = 2.1, y = 0.45, xend = 1.80, yend = .42, 
           arrow = arrow(length = unit(2, "mm")) , color = colors[2]) +
  annotate(geom = "text", x = 2.125, y = 0.45, label = "Invertebrates", 
           hjust = "left", color = colors[2]) + 
  
    annotate(geom = "segment", x = 2.1, y = 0.05, xend = 1.80, yend = .1, 
           arrow = arrow(length = unit(2, "mm")) , color = colors[3]) +
  annotate(geom = "text", x = 2.125, y = 0.05, label = "Other", 
           hjust = "left", color = colors[3]) +
  theme(legend.position="none") +
  ggtitle("Figure 6.1")
  
```


### 6.1.4 Checking multinomial Model Goodness of Fit

$$G^2 = 2\sum \mathrm{observed[log(observed/fitted)]}.$$

### 6.1.5 Example: Belief in Afterlife

```{r 6.1.5a}
Afterlife <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Afterlife.dat",
                        header = TRUE, stringsAsFactors = TRUE)
Afterlife

library(VGAM)
fit <- vglm(cbind(yes, undecided, no) ~ gender + race, family = multinomial,
            data = Afterlife)

summary(fit)
```

```{r 6.1.5b}
fit.gender <- vglm(cbind(yes, undecided, no) ~ gender, family = multinomial,
            data = Afterlife)

deviance(fit.gender)
VGAM::lrtest(fit, fit.gender) # deviance dif = 2(log-likelihood. diff.)
```

```{r 6.1.5c}
fit.race <- vglm(cbind(yes, undecided, no) ~ race, family = multinomial,
            data = Afterlife)

deviance(fit.race)
VGAM::lrtest(fit, fit.race) # deviance dif = 2(log-likelihood. diff.)
```

```{r 6.1.5d}
data.frame(Afterlife$race, Afterlife$gender, round(fitted(fit), 4))
```

### 6.1.6 Discrete Choice Models*

### 6.1.7 Example: Shopping Destination Choice*

## 6.2 Cumulative Logit Models for Ordinal Responses {x6.2}

### 6.2.1 Cumulative Logit Models with Proportional Odds {#x6.2.1}

\begin{equation} 
  \mathrm{logit}[P(Y \le j)] = \alpha_j + \beta x,\ j = 1, \dots,\ c-1
  (\#eq:eq64)
\end{equation} 

```{r}
# https://rpubs.com/riazakhan94/logstcdistbasics
cdf=function(x,mu,s){
  k=(x-mu)/s
  return(1/(1+exp(-k)))
}

theData <- tibble(x=seq(-10,12,0.01)) %>% 
  mutate(curve0 = cdf(x, 0, 1)) %>% 
  mutate(curve2 = cdf(x, 2, 1)) %>% 
  mutate(curve4 = cdf(x, 4, 1))
  
library(ggthemes)  # theme_few 
theData %>% 
  ggplot(aes(x = x)) +
  geom_line(aes(y = curve0), color = colors[3]) +
  geom_line(aes(y = curve2), color = colors[2]) +
  geom_line(aes(y = curve4), color = colors[1]) +
  ggtitle("Figure 6.2") + 
  ylab(expression("P(Y" <= "j)")) +
  scale_y_continuous(breaks=c(0,1),
        labels=c("0", "1")) +
  theme_few() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
    annotate(geom = "segment", x =-5, y = 0.55, xend = 0.0, yend = .55, 
           arrow = arrow(length = unit(2, "mm")), color = colors[3]) +
    annotate(geom = "text", x = -8, y = 0.55, label = paste("P(Y <= 3)"), 
             parse =  TRUE,hjust = "left", color = colors[3])  +

    annotate(geom = "segment", x =5, y = 0.65, xend = 2.75, yend = .65, 
           arrow = arrow(length = unit(2, "mm")) , color = colors[2]) +
    annotate(geom = "text", x = 5, y = 0.65, label = paste("P(Y <= 2)"), 
             parse = TRUE, hjust = "left", color = colors[2]) +


    annotate(geom = "segment", x =7.5, y = 0.75, xend = 5.25, yend = .75, 
           arrow = arrow(length = unit(2, "mm")) , color = colors[1]) +
    annotate(geom = "text", x = 7.5, y = 0.75,  label = paste("P(Y <= 1)"), 
             parse = TRUE, hjust = "left", color = colors[1]) 
```

$$\frac{P(Y \le j\ |\ x = a) / P(Y > j\ |\ x = a)}{P(Y \le j\ |\ x = b) / P(Y > j\ |\ x = b)}.$$

\begin{equation} 
  \mathrm{logit}[P(Y \le j)] = \alpha_j + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p,\ j = 1, \dots,\ c-1
  (\#eq:eq65)
\end{equation} 

### 6.2.2 Example: Political Ideology and Political Party Affiliation {#x6.2.2}

```{r 6.2.2a}
Polviews <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Polviews.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Polviews

library(VGAM)
# parallel = TRUE imposes proportional odds structure
# 4 intercepts for 5 y categories 
fit <-  vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender, 
             family = cumulative(parallel = TRUE),
             data = Polviews)
summary(fit)  # same effects for all 4 logits
```

$$P(Y \le j) = \frac{\mathrm{exp}(\alpha_j + \beta_1 x_1 + \beta_2 x_2)}{1 + \mathrm{exp}(\alpha_j + \beta_1 x_1 + \beta_2 x_2)}.$$

$$P(Y=3)= P(Y \le 3) - P(Y \le 2).$$


```{r 6.2.2b}
Polviews <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Polviews.dat",
                        header = TRUE, stringsAsFactors = TRUE)

# y1 = very lib, y5 = very conservative.
data.frame(Polviews$gender, Polviews$party, round(fitted(fit), 4))
```

### 6.2.3 Inference about Cumulative Logit Model Parameters

```{r 6.2.3}
fit2 <-  vglm(cbind(y1,y2,y3,y4,y5) ~ gender, 
             family = cumulative(parallel = TRUE),
             data = Polviews)
VGAM::lrtest(fit, fit2)

fit <-  vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender, 
             family = cumulative(parallel = TRUE),
             data = Polviews)

# profile likelihood CIs for beta_1 and beta_2 in full model
confint(fit, method = "profile")  

fit <-  vglm(cbind(y5,y4,y3,y2,y1) ~ party + gender, 
             family = cumulative(parallel = TRUE),
             data = Polviews)

library(tidyverse)
confint(fit, method = "profile")  %>% 
  data.frame() %>% 
  rename(LCL= X2.5..) %>% 
  rename(UCL= X97.5..) %>% 
  rownames_to_column(var = "Param") %>% 
  filter(Param == "partyrepub") %>% 
  mutate(`LCL OR` = exp(LCL)) %>% 
  mutate(`UCL OR` = exp(UCL)) %>% 
  mutate(across(c("LCL", "UCL"), round, 3)) %>% 
  mutate(across(c("LCL OR", "UCL OR"), round, 1))
```

### 6.2.4 Increased Power for Ordinal Analyses

### 6.2.5 Example: Happiness and Family Income

```{r 6.2.5a}
Happy <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Happy.dat",
                    header = TRUE, stringsAsFactors = TRUE)

Happy

fit <- vglm(cbind(y1, y2, y3) ~ income, family = cumulative(parallel = TRUE),
            data = Happy)
fit0 <- vglm(cbind(y1, y2, y3) ~ 1, family = cumulative(parallel = TRUE),
            data = Happy)
VGAM::lrtest(fit, fit0)
```

```{r 6.2.5b}
fit2 <- vglm(cbind(y1, y2, y3) ~ factor(income), family = multinomial,
            data = Happy)
fit0 <- vglm(cbind(y1, y2, y3) ~ 1, family = multinomial,
            data = Happy)
# baseline cat logit null model equivalent to cumulative logit null model
VGAM::lrtest(fit2, fit0)
```

### 6.2.6 Latent Variable Linear Models Imply Cumulative Link Models {#x6.2.6}

$$y = j\ \mathrm{if}\ \alpha_{j-1} < y^* \le \alpha_j.$$

$$Y^* = \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p + \epsilon,$$

\begin{equation} 
  \mathrm{link}[P(Y \le j)] = \alpha_j - \beta_1 x_1 - \beta_2 x_2 - \cdots - \beta_p x_p,\ j = 1, \dots,\ c-1
  (\#eq:eq66)
\end{equation} 

### 6.2.7 Invariance to Choice of Response Categories

## 6.3 Cumulative Link Models: Model Checking and Extensions *

### 6.3.1 Checking Ordinal Model Goodness of Fit

```{r 6.3.1}
Polviews <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Polviews.dat",
                        header = TRUE, stringsAsFactors = TRUE)

library(VGAM)
# parallel = TRUE imposes proportional odds structure
# 4 intercepts for 5 y categories 
fit <-  vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender, 
             family = cumulative(parallel = TRUE),
             data = Polviews)

fitWithInteraction <-  
  vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender + party:gender, 
       family = cumulative(parallel = TRUE),
       data = Polviews)

VGAM::lrtest(fitWithInteraction, fit)

```

### 6.3.2 Cumulative Logit Model without Proportional Odds

```{r}
notParallel <- summary(vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender, 
                            family = cumulative, # parallel=FALSE by default
                            data = Polviews))

parallel <- summary(vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender, 
             family = cumulative(parallel=TRUE),
             data = Polviews))

VGAM::lrtest(notParallel, parallel)
```

### 6.3.3 Simpler Interpretatoins Using Probabiliteis

### 6.3.4 Example: Modeling Mental Impairment

$$\mathrm{logit}[P(Y \le j)] = \alpha_j - \beta_1 x_1 - \beta_2 x_2,$$

```{r}

Mental <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Mental.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Mental %>% 
  filter(row_number() %in% c(1, 2, n())) 

# polr() function requires response to be a factor
y <- factor(Mental$impair)

fit <- MASS::polr(y ~ life + ses, method = "logistic", data = Mental)
summary(fit)
```

```{r}
predict(fit, data.frame(ses = 0, life = mean(Mental$life)), type = "probs")
predict(fit, data.frame(ses = 1, life = mean(Mental$life)), type = "probs")
```

```{r}
predict(fit, data.frame(ses = 0, life = min(Mental$life)), type = "probs")
predict(fit, data.frame(ses = 0, life = max(Mental$life)), type = "probs")

predict(fit, data.frame(ses = 1, life = min(Mental$life)), type = "probs")
predict(fit, data.frame(ses = 1, life = max(Mental$life)), type = "probs")
```

```{r}
source("./ocAME.R")
ocAME(fit)
```


### 6.3.5 A Latent Variable Probability Comparison of Groups

$$\hat P(Y_2^* > Y_1^*) = \mathrm{exp}(\hat\beta_2/\sqrt{2})/[1 + \mathrm{exp}(\hat\beta_2/\sqrt{2})] = 0.31$$

### 6.3.6 Cumulative Probit Model

```{r 6.3.6}
Polviews2 <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Polviews2.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Polviews2 %>% 
  filter(row_number() %in% c(1, 2, n())) 

y <- factor(Polviews2$ideology)

fit.probit <- MASS::polr(y ~ party + gender,
                         method = "probit",
                         data = Polviews2)
summary(fit.probit)  # same effects for all 4 logits
```

### 6.3.7 $R^2$ Based on Latent Variable Model

$$R_L^2 = \frac{\sum_i(y_i^* - \bar y^*)^2 - \sum_i(\hat y_i^* - \bar y^*)^2}{\sum_i(y_i^* - \bar y^*)^2} = \frac{\sum_i(\hat y_i^* - \bar y^*)^2}{\sum_i(y_i^* - \bar y^*)^2},$$

```{r 6.3.7}
fit.logit <- MASS::polr(y ~ party + gender, method = "logistic", 
                        data = Polviews2)

# lp = linear predictor
# R-squared based on logistic latent variable model
r <- var(fit.logit$lp)/ (var(fit.logit$lp) + pi^2/3)
r
sqrt(r)
var(fit.probit$lp) / (var(fit.probit$lp) + 1.0)
```

### 6.3.8 Bayesian Inference for Multinomial Models

### 6.3.9 Example: Modeling Mental Impairment Revisited

```{r 6.3.9}

Mental2 <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Mental2.dat",
                        header = TRUE, stringsAsFactors = TRUE)
Mental2 %>% 
  filter(row_number() %in% c(1, 2, n())) 

fit_freq <- MASS::polr(impair ~ life + ses, method = "logistic", data = Mental2)
summary(fit_freq)
```

```{r eval=FALSE}
#install_github("tjmckinley/BayesOrd")
library(BayesOrd)

# mnb, varb are mean, var. of beta's; vart = var of intercepts
fit <- bayesord(impair ~ life + ses, fixed = TRUE, mnb = 0, varb = 100,
                niter = 1e+6, nchains = 2, start = 10000, 
                data = Mental2)

summary(fit, digits = 3)

props <- as.matrix(fit$beta)

# posterior P(beta > 0) for each effect parameter
apply(props, 2, function(x)sum(x> 0)/length(x))
```


## 6.4 Paired-Category Logit Modeling of Ordinal Response *

### 6.4.1 Adjacent-Categories Logits

$$\mathrm{log}\left(\frac{\pi_j}{\pi_{j+1}}\right),\ j = 1, \dots, c-1.$$
\begin{equation} 
  \mathrm{log}\left(\frac{\pi_j}{\pi_{j+1}}\right) = \alpha_j + \beta x,\ j = 1, \dots, c-1.
  (\#eq:eq67)
\end{equation} 

### 6.4.2 Example: Political Ideology Revisited

```{r}
fit <- vglm(cbind(y1,y2,y3,y4,y5) ~ party + gender, 
           family = acat(parallel = TRUE, reverse = TRUE), 
           data = Polviews)
# family=acat gives adjacent-category logits
summary(fit)
```

```{r}
pValue <- 1 - pchisq(13.47, 10)
format(round(pValue, 8), scientific = FALSE)
```

### 6.4.3 Sequential Logits

### 6.4.4 Example: Tonsil Size and Streptococcus

```{r}
Tonsils <- tibble(carrier = c("yes", "no"), y1 = c(19, 497), 
                  y2 = c(29, 560), y3 = c(24, 269))

fit <- vglm(cbind(y1, y2, y3) ~ carrier, family = sratio(parallel = TRUE),
            data = Tonsils)
summary(fit)
```
