# 7 Loglinear Models for Contingency Tables and Counts {#Loglinear}

## 7.1 Loglinear Models for Counts in Contingency Tables

$$\pi_{ij} = P(X=i)P(Y=j) = \pi_{i+}\pi_{+j},\ i = 1, \dots, r,\ j = 1, \dots, c.$$

### 7.1.1 Loglinear Model of Independence for Two-Way Contingency Tables

\begin{equation} 
  P(Y=1) = \mathrm{log}\mu_{ij} = \lambda + \lambda_i^X + \lambda_j^Y,
  (\#eq:eq71)
\end{equation} 

### 7.1.2 Interpretation of Parameters in the Independence Model

### 7.1.3 Example: Happiness and Belief in Heaven

```{r}
HappyHeaven <-  
  read.table("http://users.stat.ufl.edu/~aa/cat/data/HappyHeaven.dat",
                        header = TRUE, stringsAsFactors = TRUE)
HappyHeaven


with(HappyHeaven, 
  questionr::wtd.table(happy, heaven, weight = count)
)

# canonical link for Poisson is log, so "(link = log)" is not necessary
# loglm() function in MASS library also fits loglinear models
fit <- glm(count ~ happy + heaven, family = poisson, data = HappyHeaven)

summary(fit)
```

### 7.1.4 Saturated Model for Two-Way Contingency Tables

$$
\mathrm{log}\mu_{ij} = \lambda + \lambda_i^X + \lambda_j^Y + \lambda_{ij}^{XY}
$$
$$\mathrm{log}\theta = \mathrm{log}\big( \frac{\mu_{11}\mu_{22}}{\mu_{12}\mu_{21}}\big) = \mathrm{log}\mu_{11} + \mathrm{log}\mu_{22} - \mathrm{log}\mu_{12} - \mathrm{log}\mu_{21}.$$  

### 7.1.5 Loglinear Models for Three-Way Contingency Tables

$$\mathrm{log}\mu_{ijk} =\lambda + \lambda_i^X+ \lambda_j^Y + \lambda_k^Z +\lambda_{ik}^{XZ} + +\lambda_{jk}^{YZ}.$$
$$\mathrm{log}\mu_{ijk} =\lambda + \lambda_i^X+ \lambda_j^Y + \lambda_k^Z + \lambda_{ij}^{XY} +\lambda_{ik}^{XZ} + +\lambda_{jk}^{YZ}.$$

### 7.1.6 Two-Factor Parameters Describe Conditional Associations

###  7.1.7 Example: Student Alcohol, Cigarette, and Marijuana Use

```{r 7-1-7table1}
Drugs <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Substance.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Drugs <- Drugs %>% 
  rename(A = "alcohol") %>% 
  rename(C = "cigarettes") %>% 
  rename(M = "marijuana")

`Table 7.1` <-  bind_cols(`Alcohol Use` = c("Yes", "", "No", ""),
  `Defendants' Race` = rep(c("Yes", "No"),2),
  matrix(Drugs$count, ncol = 2,byrow = T,
         dimnames = list(NULL,
         c("Marijuana Use (Yes)", "Marijuana Use (No)")))) 

knitr::kable(`Table 7.1`)
```

```{r 7-1-7table2}
A_C_M <- glm(count ~ A + C + M, family = poisson, data = Drugs)
`(A, C, M)` <- round(exp(predict(A_C_M, data.frame(Drugs))), 1) 

AM_CM <- glm(count ~ A + C + M + A:M + C:M, family = poisson, data = Drugs)
`(AM, CM)` <- round(exp(predict(AM_CM, data.frame(Drugs))), 2) 

AC_AM_CM <- glm(count ~ A + C + M + A:C + A:M + C:M, family = poisson, data = Drugs)
`(AC, AM, CM)` <- round(exp(predict(AC_AM_CM, data.frame(Drugs))), 1) 

ACM <- glm(count ~ A + C + M + A:C + A:C + A:M + C:M + A:C:M, family = poisson, data = Drugs)
`(ACM)` <- round(exp(predict(ACM, data.frame(Drugs))), 1)

`Table 7.2` <-  bind_cols(`Alcohol Use` = c("Yes", rep("", 3), "No", rep("",3)),
  `Cigarette Use` = rep(c("Yes", " ", "No", ""),2),
  `Marijuana Use` = rep(c("Yes", "No"), 4),
  `(A, C, M)` = `(A, C, M)`,
  `(AM, CM)` = `(AM, CM)`,
  `(AC, AM, CM)` = `(AC, AM, CM)`,
  `(ACM)` = `(ACM)`) 

knitr::kable(`Table 7.2`)
```

```{r 7-1-7table3}
# Table 7.3
line2 <- round(exp(coef(AM_CM)[c("Ayes:Myes", "Cyes:Myes")]), 1)
line3 <- round(exp(coef(AC_AM_CM)[c("Ayes:Cyes", "Ayes:Myes", "Cyes:Myes")]), 1)
line4 <- round(exp(coef(ACM)[c("Ayes:Cyes", "Ayes:Myes", "Cyes:Myes")]), 1)

`Table 7.3` <-  
  bind_cols(Model = c("(A, C, M)",  "(AM, CM)", "(AC, AM, CM)", "(ACM)"),
            matrix(c(rep (1.0, 3), 
                     1.0, line2["Ayes:Myes"], line2["Cyes:Myes"],
                     line3["Ayes:Cyes"], line3["Ayes:Myes"], line3["Cyes:Myes"],
                     line4["Ayes:Cyes"], line4["Ayes:Myes"], line4["Cyes:Myes"]), 
                   ncol = 3, byrow = T,
                   dimnames = list(NULL, c("AC", "AM", "CM"))))

knitr::kable(`Table 7.3`)

```

$$2.7 = \frac{(909.24 + 438.84)*(0.24 + 179.84)}{(45.76 + 555.16)*(4.76 + 142.16)}.$$

```{r, eval=FALSE}
AM_CM <- glm(count ~ A + C + M + A:M + C:M, family = poisson, data = Drugs)
round(exp(predict(AM_CM, data.frame(Drugs))), 2)  # Table 7.2 
round(exp(coef(AM_CM)), 1) # Table 7.3

is2.7 <- ((909.2395833 + 438.8404255)*(0.2395833 + 179.8404255)) / 
  ((45.7604167 + 555.1595745)*(4.7604167 + 142.1595745))

# collapse over M
AC <-
  Drugs %>% 
  group_by(A, C) %>% 
  summarise(Count2 = sum(count), .groups = "drop_last")

AC_marginal <- glm(Count2 ~ A + C + A:C, family = poisson, data = AC)
round(exp(predict(AC_marginal, data.frame(AC))), 2)
round(exp(coef(AC_marginal)), 5)

(1449*281)/(500*46)
```


```{r 7-1-7}
Drugs <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Substance.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Drugs %>% 
  filter(row_number() %in% c(1, n()))

Drugs <- Drugs %>% 
  rename(A = "alcohol") %>% 
  rename(C = "cigarettes") %>% 
  rename(M = "marijuana")

#A <- Drugs$alcohol
#C <- Drugs$cigarettes
#M <- Drugs$marijuana


fit <- glm(count ~ A + C + M + A:C + A:M + C:M, family = poisson, data = Drugs)
summary(fit)
```

## 7.2 Statistical Inference for Loglinear Models

### 7.2.1 Chi-Squared Goodness-of-Fit Tests

```{r table7-4}
# p-value for the (AC, AM, CM) model

AC_AM <- glm(count ~ A + C + M + A:C + A:M, family = poisson, data = Drugs)
AC_CM <- glm(count ~ A + C + M + A:C + C:M, family = poisson, data = Drugs)
AM_CM <- glm(count ~ A + C + M + A:M + C:M, family = poisson, data = Drugs)
AC_AM_CM <- glm(count ~ A + C + M + A:C + A:M + C:M, family = poisson, data = Drugs)

# function to convert to a p-value and return "< 0.0..." with a digit threshold 
residualP <- function(x, digits = 2){
  pValue <- 1 - pchisq(deviance(x), df.residual(x))
  value <- format(round(pValue, digits), scientific = FALSE)
  if(value == 0) {
    paste0("< 0.", strrep("0", digits-1), "1")
    } else {
      value
    }
}

`Table 7.4` <-  
  bind_cols(Model = c("(AC, AM)", "(AC, CM)", "(AM, CM)", "(AC, AM, CM)"),
            Deviance = c(deviance(AC_AM), deviance(AC_CM), deviance(AM_CM), deviance(AC_AM_CM)),
            df = c(df.residual(AC_AM), df.residual(AC_CM), df.residual(AC_AM), df.residual(AC_AM_CM)),
            `P-value` = c(residualP(AC_AM, 4), residualP(AC_CM, 4), residualP(AC_AM, 4), residualP(AC_AM_CM))) %>% 
  mutate(Deviance = round(Deviance, 1))
              

kable(`Table 7.4`)
```

### 7.2.2 Cell Standardized Residuals for Loglinear Models {#x7.2.2}

```{r}
fit <- glm(count ~ A + C + M + A:C + A:M + C:M, family = poisson, data = Drugs)
fit2 <- glm(count ~ A + C + M + A:M + C:M, family = poisson, data = Drugs)
deviance(fit)
deviance(fit2)

res <- round(rstandard(fit, type = "pearson"), 3)
res2 <- round(rstandard(fit2, type = "pearson"), 3)

tibble(Alcohol = Drugs$A, Cigarettes = Drugs$C, Marijuana = Drugs$M, 
           Count = Drugs$count, 
           "Fitted from fit" = fitted(fit), 
           "Std. Resid. from fit" = rstandard(fit, type = "pearson"),
           "Fitted from fi2" = fitted(fit2), 
           "Std. Resid. from fit2" = rstandard(fit2, type = "pearson")) %>% 
  mutate(across(contains("fit"), round, 3))
```


### 7.2.3 Significance Tests about Conditional Associations

```{r}
library(car)
Anova(fit)
```


### 7.2.4 Confidence Intervals for Conditional Odds Ratios

```{r}
fit <- glm(count ~ A + C + M + A:C + A:M + C:M, family = poisson, data = Drugs)
summary(fit)

exp(confint(fit))
```

> Caution: Notice the LCL is in scientific notation and UCL is not.

### 7.2.5 Bayesian Fitting of Loglinear Models

```{r}
#  library(MCMCpack)

fitBayes <- MCMCpack::MCMCpoisson(count ~ A + C + M + A:C + A:M + C:M, 
                                  family = poisson, data = Drugs)
summary(fitBayes)

# posterior prob. that AM log odds ratio < 0 
# (parameter 6 in model is AM log odds ratio)
mean(fitBayes[, 6] < 0)
```

### 7.2.6 Loglinear Models for Higher-Dimensional Contingency Tables


### 7.2.7 Example: Automobile Accidents and Seat Belts

### 7.2.8 Interpreting Three-Factor Interaction Terms

```{r}
Accidents <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Accidents2.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Accidents %>% 
  filter(row_number() %in% c(1, n())) 

Accidents <- 
  Accidents %>% 
  rename("G" = gender, "L" = location, "S" = seatbelt, "I" = injury)

# G*I = G + I + G:I
fit <- glm(count ~ G*L*S + G*I + L*I + S*I, family = poisson, data = Accidents)
summary(fit)
```

### 7.2.9 Statistical Versus Practical Significance: Dissimilarity Index

$$D = \sum |n_i - \hat\mu_i|/2n = \sum |p_i - \hat\pi_i|/2.$$

```{r}
fit <- glm(count ~ G*L*S + G*I + L*I + S*I, family = poisson, data = Accidents)
# dissimilarity index for loglinear model (GLS, GI, LI, SI)

DI <- sum(abs(Accidents$count - fitted(fit)))/(2*sum(Accidents$count))
round(DI, 5)

fit2 <- glm(count ~ G*L + G*S + G*I + L*S + L*I + S*I, family = poisson, 
                   data = Accidents)
# dissimilarity index for loglinear model (GLS, GI, LI, SI)
DI2 <- sum(abs(Accidents$count - fitted(fit2)))/(2*sum(Accidents$count))
round(DI2, 5)
```

## 7.3 The Loglinear - Logistic Model Connection

### 7.3.1 Using Logistic Models to Interpret Loglinear Models

### 7.3.2 Example: Auto Accident Data Revisited

\begin{equation} 
  \mathrm{logit}[P(I=1)]=\alpha + \beta_g^G + \beta_l^L + \beta_s^S.
  (\#eq:eq72)
\end{equation} 

```{r}
Injury <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Injury_binom.dat",
                        header = TRUE, stringsAsFactors = TRUE)

# 8 lines in data file, one for each binomial on injury given (G, L, S)
Injury %>% 
  filter(row_number() %in% c(1, 2, n())) 

Injury <- Injury %>% 
  rename("G" = gender,
         "L" = location,
         "S" = seatbelt)

fit2 <- glm(yes/(no + yes) ~ G + L + S, family = binomial, weights = no+yes, 
            data = Injury)

summary(fit2)
```

### 7.3.3 Condition for Equivalent Loglinear and Logistic Models

### 7.3.4 Loglinear/Logistic Model Selection Issues

## 7.4 Independence Graphs and Collapsibility

### 7.4.1 Independence Graphs

```{r}

library(igraph)
# pairs of vertices to connect 
g <- graph(c("W","X",  "Y","Z",  "Y","W",  "Z","W"), directed = FALSE)

LO <- layout_nicely(g) # original layout
angle <- 2*pi * 7.495/12  # amount of clock face to rotate
RotMat <- matrix(c(cos(angle),sin(angle),-sin(angle), cos(angle)), ncol=2)
LO2 <- LO %*% RotMat

plot(g, vertex.shape = "none", layout = LO2)
```

```{r, eval=FALSE}
# Manually draw the plot
tibble(x = c(0, 1, 2, 2, 1), 
     y = c(0, 0, 1, -1, 0), 
     name = c("X", "W", "Y", "Z", "W") ) %>% 
ggplot(aes(x = x, y = y)) +
geom_path() +
geom_point(size = 8, color = "white") +
theme_void() +
geom_text(aes(label=name), hjust= .5, vjust= .4)
```

```{r, fig.height=1.5}
tibble(x = c(0, 1, 2, 3), 
       y = c(0, 0, 0, 0), 
       name = c("W", "X", "Y", "Z") ) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_path() +
  geom_point(size = 8, color = "white") +
  theme_void() +
  geom_text(aes(label=name), hjust= .5, vjust= .4)
```

### 7.4.2 Collapsibility Conditions for Contingency Tables

```{r, fig.height=1.5}

tibble(x = c(0, 1, 2), 
       y = c(0, 0, 0), 
       name = c("A", "M", "C") ) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_path() +
  geom_point(size = 8, color = "white") +
  theme_void() +
  geom_text(aes(label=name), hjust= .5, vjust= .4)
```

```{r fig.height=1.5}
tibble(x = c(0, 1, 2), 
       y = c(0, 0, 0), 
       name = c("A", "B", "C") ) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_path() +
  geom_point(size = 8, color = "white") +
  theme_void() +
  geom_text(aes(label=name), hjust= .5, vjust= .4)
```

### 7.4.3 Example: Loglinear Model Building for Student Substance Use

```{r}
Drugs2 <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Substance2.dat",
                        header = TRUE, stringsAsFactors = TRUE)
library(dplyr)

`Table 7.5` <-  bind_cols(`Alcohol Use` = c("Yes", "", "No", ""),
  `Cigarette Use` = rep(c("Yes", "No"),2),
  matrix(Drugs2$count, ncol = 8,byrow = T,
         dimnames = list(NULL,
         c("Y_F_W", "N_F_W", "Y_M_W", "N_M_W", # Yes/No Female/Male whites
           "Y_F_O", "N_F_O", "Y_M_O", "N_M_O"))))  # whites

#knitr::kable(`Table 7.5`)


library(flextable)

my_header <- data.frame(
  col_keys= colnames(`Table 7.5`),
  line1 = c("Alcohol Use", "Cigarette Use", rep("Marijuna Use", 8)),
  line2 = c("Alcohol Use", "Cigarette Use", rep("White", 4), rep("Other", 4)),
  line3 = c("Alcohol Use", "Cigarette Use", rep(c(rep("Female", 2), rep("Male", 2)),2)),
  line4 = c("Alcohol Use", "Cigarette Use", rep(c("Yes", "No"), 4))
  )

flextable(`Table 7.5`) %>% 
  set_header_df(
    mapping = my_header,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all")

```

```{r}

Drugs2 <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Substance2.dat",
                        header = TRUE, stringsAsFactors = TRUE)

#A <- Drugs$alcohol
#C <- Drugs$cigarettes
#M <- Drugs$marijuana


fit1 <- glm(count ~ A + C + M + R + G + G:R, family = poisson, data = Drugs2)
fit2 <- glm(count ~ (A + C + M + R + G)^2, family = poisson, data = Drugs2)
fit3 <- glm(count ~ (A + C + M + R + G)^3, family = poisson, data = Drugs2)
fit4 <- glm(count ~ A + C + M + R + G + A:C + A:M + C:G + C:M + A:G + A:R + G:M + G:R + M:R, family = poisson, data = Drugs2)
fit5 <- glm(count ~ A + C + M + R + G + A:C + A:M +       C:M + A:G + A:R + G:M + G:R + M:R, family = poisson, data = Drugs2)
fit6 <- glm(count ~ A + C + M + R + G + A:C + A:M +       C:M + A:G + A:R + G:M + G:R, family = poisson, data = Drugs2)
# summary(fit1)
# summary(fit2)
# summary(fit3)
# summary(fit4)
# summary(fit5)
# summary(fit6)

`Table 7.9` <- tibble(Model = c("1. Mutual independence + GR", 
                                "2. Homogeneous association", 
                                "3. All three-factor terms",
                                "4. AC, AM, CG, CM, AG, AR, GM, GR, MR",
                                "5. AC, AM,     CM, AG, AR, GM, GR, MR",
                                "6. AC, AM,     CM, AG, AR, GM, GR"),
                      Deviance = c(deviance(fit1), deviance(fit2), 
                                   deviance(fit3), deviance(fit4),
                                   deviance(fit5), deviance(fit6)),
                      df = c(df.residual(fit1), df.residual(fit2), 
                             df.residual(fit3), df.residual(fit4),
                             df.residual(fit5), df.residual(fit6))) %>% 
  mutate(Deviance = round(Deviance, 2))

`Table 7.9`
```

### 7.4.4 Collapsibility and Logistic Models

## 7.5 Modeling Ordinal Associations in Contingency Tables

```{r}
Teenagers <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Teenagers.dat",
                        header = TRUE, stringsAsFactors = TRUE) %>% 
  mutate(`Premarital Sex` = recode_factor(sex, "1" = "Always wrong", 
                                   "2" = "Almost always wrong",
                                   "3" = "Wrong only sometimes",
                                   "4" = "Not wrong at all")) %>% 
  mutate(`Teenage Birth Control` = recode_factor(birth, "1" = "Strongly Disagree", 
                                          "2" = "Disagree",
                                          "3" = "Agree",
                                          "4" = "Strongly Agree"))

fit <- glm(count ~ factor(sex) + factor(birth), family=poisson, 
           data = Teenagers)
#summary(fit)  # shows Residual deviance: 127.65  on  9  degrees of freedom
#fitted(fit)

fitLinear <- glm(count ~ factor(sex) + factor(birth) + sex:birth, family=poisson, 
           data = Teenagers)
# summary(fitLinear)  # Residual deviance:  11.534  on  8  degrees of freedom
#fitted(fitLinear)
# raw data 
# xtabs(Teenagers$count ~ Teenagers$sex + Teenagers$birth)

with(Teenagers, xtabs(count ~ `Premarital Sex` + `Teenage Birth Control`))

TeenagersPlus <- bind_cols(Teenagers, independence = round(fitted(fit), 1), 
                           linear = round(fitted(fitLinear), 1))

with(TeenagersPlus, xtabs(independence ~ `Premarital Sex` + `Teenage Birth Control`))

with(TeenagersPlus, xtabs(linear ~ `Premarital Sex` + `Teenage Birth Control`))

```

### 7.5.1 Linear-by-Linear Association Model

$$\mathrm{log}\mu_{ij}= \lambda + \lambda_i^X + \lambda_j^Y + \beta\upsilon_i\nu_j.$$
\begin{equation} 
  \frac{\mu_{ab}\mu_{cd}}{\mu_{ad}\mu_{cg}} = \mathrm{exp}[\beta(\upsilon_c - \upsilon_a)(\nu_d - \nu_b)] 
  (\#eq:eq73)
\end{equation} 

### 7.5.2 Example: Linear-by-Linear Association for Sex Opinions

```{r}
Teenagers <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Teenagers.dat",
                        header = TRUE, stringsAsFactors = TRUE)
Teenagers %>% 
  filter(row_number() %in% c(1, 2, n())) 

# quantitative sex-by birth interaction term
fit <- glm(count ~ factor(sex) + factor(birth) + sex:birth, family=poisson, 
           data = Teenagers)
summary(fit)

# interaction is likelihood ratio test for L x L association term
car::Anova(fit)
```

$$\mathrm{exp}[\beta(\upsilon_4 - \upsilon_1)(\nu_4 - \nu_1)] = \mathrm{exp}[0.286(4-1)(4-1) = 13.1$$

### 7.5.3 Ordinal Significance Tests of Independence

## 7.6 Loglinear Modeling of Count Response Variables *

### 7.6.1 Count Regressoin Modeling of Rate Data

$$\mathrm{log}(\mu/t)= \alpha + \beta_1 x_1 + \cdots + \beta_p x_p.$$

$$\mathrm{log} \mu - \mathrm{log} t = \alpha + \beta_1 x_1 + \cdots + \beta_p x_p.$$

$$\mu = t\ \mathrm{exp}(\alpha + \beta_1 x_1 + \cdots + \beta_p x_p).$$

### 7.6.2 Example: Death Rates for Lung Cancer Patients

```{r}
Cancer <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Cancer.dat",
                        header = TRUE, stringsAsFactors = TRUE)

cancerCountWide <- Cancer %>% 
  select(-risktime) %>% 
  pivot_wider(id_cols = time, names_from = c(histology, stage), 
              values_from=count) %>% 
  mutate(time = case_when(time == 1 ~ "0-2",
                          time == 2 ~ "2-4",
                          time == 3 ~ "4-6",
                          time == 4 ~ "6-8",
                          time == 5 ~ "8-10",
                          time == 6 ~ "10-12",
                          time == 7 ~ "12+",)) %>% 
  mutate(`histo` = " ") %>% # blank column for column headings
  select(time, histo, everything())

cancerRiskTimeWide <- Cancer %>% 
  select(-count) %>% 
  mutate(risktime = paste0("(", risktime, ")")) %>% 
  pivot_wider(id_cols = time, names_from = c(histology, stage), 
              values_from=risktime) %>% 
  mutate(time = case_when(time == 1 ~ "0-2",
                          time == 2 ~ "2-4",
                          time == 3 ~ "4-6",
                          time == 4 ~ "6-8",
                          time == 5 ~ "8-10",
                          time == 6 ~ "10-12",
                          time == 7 ~ "12+",)) %>% 
  mutate(`histo` = " ") %>% 
  select(time, histo, everything())

# function to interleave two matrices
# https://stackoverflow.com/questions/19781723/interleave-rows-of-matrix-stored-in-a-list-in-r
# https://gist.github.com/mrdwab/7313857
Interleave <- function(myList, append.source = TRUE, sep = ": ", drop = FALSE) {
  sources <- myList
  sources[sapply(sources, is.null)] <- NULL
  sources <- lapply(sources, function(x) {
    if (is.matrix(x) || is.data.frame(x)) {
      x
    } else {
      t(x)
    }
  })
  nrows <- sapply(sources, nrow)
  mrows <- max(nrows)
  if (any(nrows != mrows & nrows != 1)) {
    stop("Arguments have differening numbers of rows.")
  }
  sources <- lapply(sources, function(x) {
    if (nrow(x) == 1) {
      x[rep(1, mrows), , drop = drop]
    } else {
      x
    }
  })
  tmp <- do.call("rbind", sources)
  nsources <- length(sources)
  indexes <- outer((0:(nsources - 1)) * mrows, 1:mrows, "+")
  retval <- tmp[indexes, , drop = drop]
  if (append.source && !is.null(names(sources))) {
    if (!is.null(row.names(tmp))) {
      row.names(retval) <- paste(format(row.names(retval)),
        format(names(sources)),
        sep = sep
      )
    } else {
      row.names(retval) <- rep(names(sources), mrows)
    }
  }
  retval
}

# objects to interleave
l <- list(a=as.matrix(cancerCountWide),b=as.matrix(cancerRiskTimeWide))

# interleave counts and risk time
bigMatrix <- Interleave(l)

# add columns for titles
biggerMatrix <- data.frame(cbind(bigMatrix[,1], bigMatrix[,2:11]))
names(biggerMatrix) <- names(cancerCountWide)

my_header <- data.frame(
  col_keys = c("time", "histo", "blank1", 
               "1_1", "2_1", "3_1", "blank2", 
               "1_2", "2_2", "3_2", "blank3", 
               "1_3", "2_3","3_3"),
  line2 = c("Follow-up", "Histology", "", rep("I", 3), "", rep("II", 3), "", 
            rep("III", 3)),
  line3 = c("Follow-up", "Disease Stage", rep(c("", "1", "2", "3"), 3))
)

library(officer)
big_border = fp_border(color="black", width = 2)

library(flextable)
flextable(biggerMatrix, col_keys = my_header$col_keys) %>% 
  set_header_df(
    mapping = my_header,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  merge_v(part = "body") %>%
  align(align = "center", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_bottom(part="body", border = big_border) %>% 
  fix_border_issues()
```

$$\mathrm{log}(\mu_{ijk}/t_{ijk}) = \beta_0 + \beta_i^H + \beta_j^S + \beta_k^T,$$

```{r}
Cancer <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Cancer.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Cancer %>% 
  filter(row_number() %in% c(1, n())) 

Cancer <- Cancer %>% 
  mutate("logrisktime" = log(Cancer$risktime))

# showing 6 time effects
fit <- glm(count ~ factor(histology) + factor(stage) + factor(time), 
           family = poisson, offset = logrisktime, data = Cancer)
summary(fit)

# likelihood-ratio test of effects, adjusting for the others
car::Anova(fit)
```

### 7.6.3 Negativee Binomial Regression Models

$$E(Y) = \mu,\ \ \mathrm{var}(Y) = \mu + D\mu^2.$$

### 7.6.4 Example: Female Horseshoe Crab Satelites Revisited

```{r}
Crabs <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Crabs.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Crabs %>% 
  filter(row_number() %in% c(1, n())) 

fit.pois <- glm(sat ~ width, family = poisson, data = Crabs)
summary(fit.pois)

fit.negbin <- MASS::glm.nb(sat ~ width, data = Crabs)
summary(fit.negbin)

```
