# 8 Models for Matched Pairs {#Matched}

```{r}
Opinions <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Envir_opinions.dat",
                        header = TRUE, stringsAsFactors = TRUE)
# Make contingency table
tab <- as.matrix(addmargins(xtabs(~y1 + y2, data = Opinions)))

library(tibble)
# Add label as the first column and variable names
`Table 8.1` <- tibble(`Pay Higher Taxes` = c("Yes", "No", "Total"), 
                      Yes = tab[,1], No = tab[,2], Total = tab[,3])

# horrible kludge
`Table 8.1`$Yes <- paste(`Table 8.1`$Yes, c(" (pi11)", " (pi21)", ""))
`Table 8.1`$No <- paste(`Table 8.1`$No, c(" (pi12)", " (pi22)", ""))

library(flextable)
my_header <- data.frame(
  col_keys = colnames(`Table 8.1`),
  line1 = c("Pay Higher Taxes", rep("Cut Living Standards", 2), "Total"),
  line2 = colnames(`Table 8.1`)
)

flextable(`Table 8.1`, col_keys = my_header$col_keys) %>%
  set_header_df(
    mapping = my_header,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  autofit(part = "all") %>%    
  align(align = "center", part = "all") %>% 
  merge_h(part = "header") %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "body") %>% 
  merge_v(part = "body") %>%
  align_nottext_col(align = "center") %>% 
  set_caption(caption = "Table 8.1") 
```


## 8.1 Comparing Dependent Proporitons for Binary Mached Pairs {#x8.1}

### 8.1.1 McNemar Test Comparing Marginal Proportions
$$H_0: P(Y_1 = 1) = P(Y_2 = 1), \mathrm{\ or\ equivalently\ } H_0: \pi_{12} = \pi_{21}.$$

\begin{equation} 
  z = \frac{n_{12}-(\frac{1}{2}n^*)}{\sqrt{n^*(\frac{1}{2})(\frac{1}{2})}}= \frac{n_{12}- n_{21}}{\sqrt{n_{12}+n_{21}}}
  (\#eq:eq81)
\end{equation} 


```{r}
Opinions <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Envir_opinions.dat",
                        header = TRUE, stringsAsFactors = TRUE)

library(dplyr)

# Data file has 1144 lines, one for each of 1144 people. 
# Each person has two binary responses (y1 and y2)
Opinions %>% 
  filter(row_number() %in% c(1, 2, n()))  

tab <- xtabs(~y1 + y2, data = Opinions)
tab

# Don't use continuity correction, which is too conservative.
mcnemar.test(tab, correct = FALSE)

library(PropCIs)

# specific to how to round 
currentDigits = unlist(options("digits"))  # save current rounding
options(digits = 1)

# 95% Wald CI for difference of marginal probabilities 
diffpropci.Wald.mp(107, 132, 1144, 0.95)  # (n21, n12, n, confidence level)
diffpropci.Wald.mp(tab[2,1], tab[1,2], sum(tab), 0.95)  

# 95% score CI for difference of marginal probabilities 
scoreci.mp(tab[2,1], tab[1,2], sum(tab), 0.95)
options(digits = currentDigits)  # reset rounding
```

### 8.1.2 Estimating the difference between Dependent Proportions

## 8.2 Marginal Models and Subject-Specific Models for Matched Pairs {#x8.2}

### 8.2.1 Marginal Models for Marginal Proportions {#x8.2.1}

$$P(Y_1 = 1) = \alpha + \delta,\ P(Y_2= 1) = \alpha,$$ 

$$P(Y_1 = 1) = \alpha + \delta x_i$$ 

\begin{equation} 
  \mathrm{logit}[P(Y_t = 1)] = \alpha + \beta x_i.
  (\#eq:eq82)
\end{equation} 

### 8.2.2 Example: Enironmental Options Revisted

```{r}
Opinions <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Opinions.dat",
                        header = TRUE, stringsAsFactors = TRUE)

# Data file has 1144 lines, one for each of 1144 people. 
# Each person has two binary responses (y1 and y2)
Opinions %>% 
  filter(row_number() %in% c(1, 2, 3, 4, n()-1, n()))  

library(gee)
# id identifies variable on which observe y1, y2
fit <- gee(y ~ question, id = person, family = binomial(link = "identity"),
           data = Opinions)

summary(fit) # question parameter for identity link is difference of proportions

fit2 <- gee(y ~ question, id = person, family = binomial(link = logit),
           data = Opinions)
summary(fit2) # question parameter for logit link is log odds ratio
```

### 8.2.3 Subject-Specific and Population-Averaging Tables {#x8.2.3}

```{r}
`Table 8.2` <- data.frame(Question = c("yi1: Pay higher taxes?", 
                                       "yi2: Cut living standards?"),
                          Yes = c(1, 1),
                          No = c(0, 0))

theHeader <- data.frame(col_keys = colnames(`Table 8.2`),
                        line1 = c("Question", rep("Response", 2)),
                        line2 = colnames(`Table 8.2`))

library(flextable)
library(dplyr)

flextable(`Table 8.2`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Table 8.2: Representation of subject-specific table for matched pair contributing to count n_11 = 227 in Table 8.1.") %>% 
  set_table_properties(width = .75, layout = "autofit")
```

### 8.2.4 Conditional Logistic Regression for Matched-Pairs* {#x8.2.4}

\begin{equation} 
  \mathrm{logit}[P(Y_{it} = 1)] = \alpha_i + \beta x_{it},\ t = 1,\ 2,
  (\#eq:eq83)
\end{equation} 


$$P(Y_{i1} = 1) = \frac{\mathrm{exp}(\alpha_i + \beta)}{1 + \mathrm{exp}(\alpha_i + \beta)},\  P(Y_{i2} = 1) = \frac{\mathrm{exp}(\alpha_i)}{1 + \mathrm{exp}(\alpha_i)}.$$  
### 8.2.5 Logistic Regression for Matched Case-Control Studies*

```{r}
library(tidyverse)
`Table 8.3` <- tibble(`Normal Birth Weight (Controls)` = 
                            c("Nonsmokers", 
                              "Smokers"),
                          Nonsmokers = c(159, 8),
                          Smokers = c(22, 14))

theHeader <- data.frame(col_keys = c("Normal Birth Weight (Controls)", "blank", "Nonsmokers", "Smokers"),
                        line1 = c("Normal Birth Weight (Controls)", "", rep("Low Birth Weight (Cases)", 2)),
                        line2 = c("Normal Birth Weight (Controls)", "blank", "Nonsmokers", "Smokers"))

library(flextable)
library(dplyr)

library(officer)
big_border = fp_border(color="black", width = 2)

flextable(`Table 8.3`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Table 8.3") %>% 
  set_table_properties(width = .75, layout = "autofit") %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_bottom(part="body", border = big_border)  
```

```{r}
library(tidyverse)
`Table 8.4` <- tibble(`Smoker` = 
                            c("No", 
                              "Yes"),
                          aCase = c(1, 0),
                          aCon = c(1, 0),
                          bCase = c(0, 1),
                          bCon = c(1, 0),
                          cCase = c(1, 0),
                          cCon = c(0, 1),
                          dCase = c(0, 1),
                          dCon = c(0, 1))

theHeader <- data.frame(col_keys = c("Smoker" ,
                         
                          "aCase",
                          "aCon", 
                          "blank1",
                          "bCase",
                          "bCon", 
                           
                          "blank2",
                          "cCase",
                          "cCon", 
                           
                          "blank3",
                          "dCase",
                          "dCon", 
                          "blank4" ),
                        line1 = c("Smoker", rep("a", 2), "", rep("b", 2), "", 
                                  rep("c", 2), "", rep("d", 2),""),
                        line2 = c("Smoker", rep(c("Case", "Control", " "), 4)))

library(flextable)
library(dplyr)

flextable(`Table 8.4`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Table 8.4") %>% 
  set_table_properties(width = .75, layout = "autofit") %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_top(part="body", border = big_border)  %>% 
  hline_bottom(part="body", border = big_border)  
```

## 8.3 Comparing Proportions for Nominal Matched-Pairs Responses

$$P(Y_1 = i) = P(Y_2 = i)\ \mathrm{for}\ i = 1, \dots, c,$$ 

### 8.3.1 Marginal Homogeneity for Baseline-Category Logit Models 

$$\mathrm{log}\left[\frac{P(Y_1 = j)}{P(Y_1 = c)}\right] = \alpha_j + \beta_j,\ \mathrm{log}\left[\frac{P(Y_2 = j)}{P(Y_2 = c)}\right] = \alpha_j$$ 
### 8.3.2 Example: Coffee Brand Market Share

```{r}
Coffee <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Coffee.dat",
                        header = TRUE, stringsAsFactors = TRUE)

library(dplyr)
library(tidyr)
coffee_wide <- Coffee %>% 
  tidyr::pivot_wider(id_cols = person, names_from=purchase, names_prefix = "y", 
                     values_from = y)
  


# Make contingency table
tab <- as.matrix(addmargins(xtabs(~y1 + y0, data = coffee_wide)))

library(tibble)
# Add label as the first column and variable names
`Table 8.5` <- tibble(`First Purchase` = c("High Point", "Taster's Choice",
                                           "Sanka", "Nescafe", "Brim", "Total"), 
                      `High Point` = tab[, 1], `Taster's Choice` = tab[,2],
                      `Sanka` = tab[, 3], `Nescafe` = tab[,4], `Brim` = tab[,5],
                      Total = tab[, 6])


theHeader <- data.frame(col_keys = c("First Purchase", "Blank", "High Point", 
                                     "Taster's Choice", "Sanka", "Nescafe", 
                                     "Brim", "Total"),
                        line1 = c("First Purchase", "", 
                                  rep("Second Purchase", 6)),
                        line2 = c("First Purchase", "", "High Point", 
                                     "Taster's Choice", "Sanka", "Nescafe", 
                                     "Brim", "Total"))

library(flextable)
library(dplyr)

library(officer)
big_border = fp_border(color="black", width = 2)

flextable(`Table 8.5`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  align(j="First Purchase", align= "left", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Table 8.5") %>% 
  set_table_properties(width = .75, layout = "autofit") %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_bottom(part="body", border = big_border)  
```
Using GEE methodology, we can fit the baseline-category logit model, as shown in the following `R` output.


```{r}
Coffee <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Coffee.dat",
                        header = TRUE, stringsAsFactors = TRUE)

# subject-specific data file, purchase is 1 for first purchase 0 for second
Coffee %>% 
  filter(row_number() %in% c(1, 2, 3, 4, n()-1, n()))  


library(multgee)  # package for multinomial GEE analysis
fit <- nomLORgee(y ~ purchase, id = person, LORstr = "independence", 
                 data = Coffee)
# san.se = robust SE
# beta01 is alpha1 in our notation
summary(fit)  # nomLORgee uses baseline category logits for nominal response 

fit0 <- nomLORgee(y ~ 1, id = person, LORstr = "independence", data = Coffee)

# Model under H_0: y ~ 1 null model (marginal homogeneity)
waldts(fit0, fit)

with(Coffee, mantelhaen.test(purchase, y, person))
```

### 8.3.3 Using Cochran-Mantel-Haenszel Test to Test marinal Homogeneity*

### 8.3.4 Symmetry and Quasi-Symmetry Models for Square Contingency Tables {#x8.3.4}

$$\pi_{ij} = \pi_{ji}$$ 
$$\mathrm{log}(\pi_{ij}/\pi_{ji}) = 0\ \mathrm{for\ all}\ i \ \mathrm{and}\ j.$$ 

$$r_{ij} = (n_{ij} - n_{ji})/\sqrt{n_{ij} + n_{ji}}.$$ 

\begin{equation} 
  \mathrm{log}(\pi_{ij}/\pi_{ji}) = \beta_i - \beta_j\  \mathrm{for\ all}\ i \ \mathrm{and}\ j.
  (\#eq:eq84)
\end{equation} 

### 8.3.5 Example: Coffee Brand market Share Revisited

```{r}
Coffee2 <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Coffee2.dat",
                        header = TRUE, stringsAsFactors = TRUE)
# quasi-symmetry model, nij and nji are opposite cell counts in table 8.5
Coffee2

symm <- glm(nij/(nij+nji) ~ -1, family = binomial, weights = nij + nji, 
            data = Coffee2)
summary(symm)  # symmetry model, -1 in model statement sets intercept = 0

QS <- glm(nij/(nij+nji) ~ -1 + H + T + S + N + B , family = binomial, 
          weights = nij + nji, data = Coffee2)
summary(QS)
```

## 8.4 Comparing Proportions for Ordinal Matched-Pairs Responses

### 8.4.1 Marginal Homogeneity and Cummulative Logit Marginal Model

$$\mathrm{logit}[P(Y_1 \le j)] = \alpha_j + \beta,\ \mathrm{logit}[P(Y_2 \le j)] = \alpha_j$$

### 8.4.2 Example: Recycle or Drive Less to Help the Environment? {#x8.4.2}

```{r}
Envir <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Envir.dat",
                        header = TRUE, stringsAsFactors = TRUE)

library(dplyr)
library(tidyr)
envir_wide <- Envir %>% 
  tidyr::pivot_wider(id_cols = person, names_from=question, names_prefix = "y", 
                     values_from = y)
  


# Make contingency table
tab <- as.matrix(xtabs(~y1 + y0, data = envir_wide))

library(tibble)
# Add label as the first column and variable names
`Table 8.6` <- tibble(`Recycle` = c("Always", "Often", "Sometimes", "Never"), 
                      `Always` = tab[, 1], `Often` = tab[,2], 
                      `Sometimes` = tab[, 3], `Never` = tab[,4])


theHeader <- data.frame(col_keys = c("Recycle", "Blank", "Always", "Often",
                                           "Sometimes", "Never"),
                        line1 = c("Recycle", "", rep("Drive Less", 4)),
                        line2 = c("Recycle", "", "Always", "Often", 
                                  "Sometimes", "Never"))

library(flextable)
library(dplyr)

library(officer)
big_border = fp_border(color="black", width = 2)

flextable(`Table 8.6`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  align(j="Recycle", align= "left", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Table 8.6") %>% 
  #set_table_properties(width = .75, layout = "autofit") %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_bottom(part="body", border = big_border)  
```

```{r}
Envir <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Envir.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Envir %>% 
  filter(row_number() %in% c(1, 2, n()-1, n()))  

library(multgee)
fit <- ordLORgee(y ~ question, id = person, LORstr = "independence", 
                 data = Envir)
summary(fit)
```

### 8.4.3 An Ordinal Quasi-Symmetry Model * 

\begin{equation} 
  \mathrm{log}(\pi_{ij}/\pi_{ji}) = \beta(u_j = u_i).
  (\#eq:eq85)
\end{equation}

### 8.4.4 Example: Recycle or Drive Less Revisited?

```{r}
Envir <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Environment.dat",
                        header = TRUE, stringsAsFactors = TRUE)

# nij and nji are opposite cell counts in Table 8.6
# x = j-1 = distance between categories, 6 pairs of opposite cells
Envir  

fit <- glm(nij/(nij+nji) ~ -1 + x, family = binomial, weight = nij + nji, 
           data = Envir)
summary(fit)  # ordinal quasi symmetry; -1 in model sets intercept = 0

fit.QS <- glm(nij/(nij+nji) ~ -1 + always + often + sometimes + never, 
              family = binomial, weight = nij + nji, data = Envir)
summary(fit.QS)  # quasi symmetry
```

## 8.5 Analyzing Rater Agreement *


### 8.5.1 Example: Agreement on Carcinoma Diagnosis

```{r}
Pathology <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Pathologists.dat",
                        header = TRUE, stringsAsFactors = TRUE)

library(dplyr)
library(tibble)

# Make contingency table
mat <- matrix(Pathology$count, ncol = 4,byrow = T) %>% 
  addmargins()
  
colnames(mat) <- c("V1", "V2", "V3", "V4", "Sum") 
  
`Table 8.7` <-  
  mat %>% 
  as_tibble() %>% 
  bind_cols(`Pathologist X` = c(1:4, "Total")) %>% 
  select(`Pathologist X`, everything())
  
theHeader <- data.frame(col_keys = c("Pathologist X", "Blank", "V1", "V2",
                                           "V3", "V4", "Sum"),
                        line1 = c("Pathologist X", "", rep("Pathologist Y", 4),
                                  "Total"),
                        line2 = c("Pathologist X", "", "1", "2", "3", "4",
                                  "Total"))

library(flextable)
library(dplyr)

library(officer)
big_border = fp_border(color="black", width = 2)

flextable(`Table 8.7`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  align(j="Pathologist X", align= "left", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Table 8.7") %>% 
  #set_table_properties(width = .75, layout = "autofit") %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_bottom(part="body", border = big_border)  
```

### 8.5.2 Cell Residuals for Independence Model

### 8.5.3 Quasi-Independence Model

```{r}
Pathology <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Pathologists.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Pathology  # diag: separate category (1, 2, 3, 4) for each main-diagonal cell 
# and a single (0) for all other cells

fit <- glm(count ~ factor(X) + factor(Y) + factor(diag), family = poisson,
           data = Pathology)

# quasi independence, not showing intercept or main effects
summary(fit)
```

### 8.5.4 Quasi Independence and Odds Ratios Summarizing Agreement

$$\tau_{ab} = \mathrm{exp}(\delta_a + \delta_b).$$ 

###  8.5.5 Kappa Summary Measure of Agreement

$$\kappa = \frac{\sum_i\pi_{ii}-\sum_i\pi_{i+}\pi_{+i}}{1-\sum_i\pi_{i+}\pi_{+i}}$$ 

```{r}
library(psych)
dat <- matrix(Pathology$count, ncol = 4, byrow = TRUE)
cohen.kappa(dat)
```

## 8.6 Bradley-Terry Model for Paired Preferences * 

```{r}
`Table 8.8` <- tibble(Winner = c("Djokovic", "Federer", "Murray", "Nadal", 
                                 "Wawrinka"),
                      Djokovic = c("-", 6, 3, 2, 3), 
                      Federer = c(9, "-", 0 ,1 , 2),
                      Murray = c(14, 5, "-", 4, 2),
                      Nadal = c(9, 5, 2, "-", 3),
                      Wawrinka = c(4, 7, 2, 4, "-"))

theHeader <- data.frame(col_keys = c("Winner", "Blank", "Djokovic", "Federer",
                                     "Murray", "Nadal", "Wawrinka"),
                        line1 = c("Winner", "", rep("Looser", 5)),
                        line2 = c("Winner", "", "Djokovic", "Federer",
                                     "Murray", "Nadal", "Wawrinka"))

library(flextable)
library(dplyr)

library(officer)
big_border = fp_border(color="black", width = 2)

flextable(`Table 8.8`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  align(j="Winner", align= "left", part = "all") %>% 
  autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Resuls of 2014-2018 matches for men tennis players.") %>% 
  #set_table_properties(width = .75, layout = "autofit") %>% 
  hline_top(part="header", border = big_border) %>% 
  hline_bottom(part="body", border = big_border)  

```


### 8.6.1 The Bradley-Terry Model of Quasi-Symmetry

$$\mathrm{logit}(\Pi_{ij}) = \mathrm{log}(\Pi_{ij}/\Pi_{ji}) = \beta_i - \beta_j.$$ 

$$\hat\Pi_{ij} = \mathrm{exp}(\hat\beta_i - \hat\beta_j)/[1 + \mathrm{exp}(\hat\beta_i - \hat\beta_j)].$$

### 8.6.2 Example: Ranking Men Tennis Players

```{r}
Tennis <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Tennis.dat",
                        header = TRUE, stringsAsFactors = TRUE)
# Djokovic won 9 lost 6 vs Fed
# Federer always beat Murray

fit <- 
  glm(nij/(nij + nji) ~ -1 + Djokovic + Federer + Murray + Nadal + Wawrinka, 
      family = binomial, weights = nij + nji, data = Tennis)
summary(fit) 
```

$$\hat\Pi_{ij} = \frac{\mathrm{exp}(\hat\beta_i - \hat\beta_j)}{1 + \mathrm{exp}(\hat\beta_i - \hat\beta_j)} = \frac{\mathrm{exp}(1.136 - 1.176)}{1 + \mathrm{exp}(1.136 - 1.176)} = 0.49$$
