# 9 Marginal Modeling of Correlated, Clustered Responses {#Marginal}

## 9.1 Marginal Models Vs Subject-Specific Models

### 9.1.1 Marginal Models for a Clustered Binary Response

### 9.1.2 Example: Repeated Responses on Similar Survey Questions {#x9.1.2}

```{r}
Abortion <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Abortion.dat",
                        header = TRUE, stringsAsFactors = TRUE)

library(tidyverse)
a_wide <- 
  Abortion %>% 
  tidyr::pivot_wider(id_cols = c(person, gender), # do not pivot
                     names_from=situation,  # variable with situation number
                     names_prefix = "sit",  # new name prefix
                     values_from = response) %>% # 
  mutate(pattern = paste(sit1, sit2, sit3, sep = ",")) %>%  # make word with pattern 
  mutate(pattern = factor(pattern, 
                          levels = c("1,1,1", "1,1,0", "0,1,1", "0,1,0", 
                                     "1,0,1", "1,0,0",  "0,0,1","0,0,0"), 
                          ordered = TRUE))

`Table 9.1` <- as.data.frame.matrix(xtabs( ~ a_wide$gender + a_wide$pattern ), 
                            row = 2 ) %>% 
  bind_cols(Gender = c("Male", "Female")) %>%  # add sex
  select(Gender, everything())  # reorder columns



theHeader <- data.frame(col_keys = colnames(`Table 9.1`),
                        line1 = c("Gender",  
                                  rep("Sequene of Responses in Three Situations", 8)),
                        line2 = colnames(`Table 9.1`))

library(flextable)

flextable(`Table 9.1`, col_keys = theHeader$col_keys) %>% 
  set_header_df(
    mapping = theHeader,
    key = "col_keys"
  ) %>% 
  theme_booktabs() %>% 
  merge_v(part = "header") %>% 
  merge_h(part = "header") %>% 
  align(align = "center", part = "all") %>% 
  align(j="Gender", align= "left", part = "all") %>% 
  #autofit() %>% 
  empty_blanks() %>% 
  fix_border_issues()  %>%
  set_caption(caption = "Support (1 = yes, 0 = no) for legalized abortion in three situations, by gender") %>% 
  set_table_properties(width = .75, layout = "autofit")
```

### 9.1.3 Subject-Specific Models for a Repeated Response

## 9.2 Marginal Modeling: The Generalized Estimating Equations (GEE) Approach

### 9.2.1 Quasi-Likelihood Methods

### 9.2.2 Generalized Estimating Equation Methodology: Basic Ideas

### 9.2.3 Example: Opinion about Legalized Abortion Revisited {#x9.2.3}

```{r}
Abortion <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Abortion.dat",
                        header = TRUE, stringsAsFactors = TRUE)
library(tidyverse)
Abortion %>% 
  filter(row_number() %in% c(1, 2, 3, n()-2, n()-1, n()))  

Abortion <- 
  Abortion %>% 
  mutate(sit = factor(situation, levels = c(3, 1, 2)))

fit.glm <- 
  glm(response ~ sit + gender, family = binomial, data = Abortion)

# ML estimate for 5550 independent observations
summary(fit.glm)
```

```{r}
library(gee)
# cluster on "id" variable
fit.gee <- gee(response ~ sit + gender, id = person, family = binomial,
               corstr = "independence", data = Abortion)
summary(fit.gee)
```

```{r}
fit.gee2 <- 
  gee(response ~ sit + gender, id = person, family = binomial,
               corstr = "exchangeable", data = Abortion)
summary(fit.gee2)


fit.geeUn <- 
  gee(response ~ sit + gender, id = person, family = binomial,
               corstr = "unstructured", data = Abortion)
summary(fit.geeUn)$working.correlation
```

### 9.2.4 Limitations of GEE Compared to ML

```{r}
library(geepack)  # geepack library enables Wald tests comparing models
fit <- 
  geeglm(response ~ gender + factor(situation) , id = person, family = binomial,
         corstr = "exchangeable", data = Abortion)
anova(fit)
```

## 9.3 Marginal Modeling for Clustered Multinomial Responses

### 9.3.1 Example: Insomnia Study {#x9.3.1}
$$\hat\beta_1 = 1.038\ (0.168),\ \hat\beta_2 = 0.034\ (0.238),\ \hat\beta_3 = 0.708\ (0.224)$$

```{r}
Insomnia <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Insomnia.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Insomnia %>% 
  filter(row_number() %in% c(1, 2, n()-1, n()))  

library(multgee)
fit <- ordLORgee(response ~ occasion + treat + occasion:treat, id = case,
                 LORstr = "independence", data = Insomnia)
summary(fit)
```

### 9.3.2 Alternative GEE Specification of Working Association

## 9.4 Transitional Modeling, Given the Past

### 9.4.1 Transitional Models with Explanatory Variables

$$\mathrm{logit}[P(Y_t = 1)] = \alpha + \beta y_{t-1} + \beta_1 x_{1t} + \cdots + \beta_p x_{pt},$$ 

### 9.4.2 Example: Respiratory Illness and Maternal Smoking
$$\mathrm{logit}[P(Y_t = 1)] = \alpha + \beta y_{t-1} + \beta_1 s + \beta_2 t,\ t = 8, 9, 10$$

```{r}
Respiratory <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Respiratory.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Respiratory 

fit <- glm(yes/(yes+no) ~ previous + s + t, family = binomial, 
           weights = no + yes, data = Respiratory)
summary(fit)

car::Anova(fit)
```

### 9.4.3 Group Comparisons Treating Initial Response as Covariate {#x9.4.3}

\begin{equation} 
\mathrm{logit}[P(Y_2 \le j)] = \alpha_j + \beta_1 x + \beta_2 y_1
  (\#eq:eq91)
\end{equation} 

```{r}
Insomnia2 <- read.table("http://users.stat.ufl.edu/~aa/cat/data/Insomnia2.dat",
                        header = TRUE, stringsAsFactors = TRUE)

Insomnia2 

library(VGAM)
fit <- vglm(cbind(follow1, follow2, follow3, follow4) ~ treatment + initial,
            family = cumulative(parallel = TRUE), data = Insomnia2)
summary(fit)

detach("package:multgee", 
       unload = TRUE)

detach("package:VGAM", 
       unload = TRUE) # contains function s that conflicts with gam
```

## 9.5 Dealing with Missing Data *

### 9.5.1 Missing at Random: Impact on ML and GEE Methods

### 9.5.2 Multiple Imputation: Monte Carlo Prediction of missing Data
